The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (8th edition -- distributed with CrypTool version 1.4.10)} \vspace{60pt}

### 2007-07-02 script/en/chapters/primes.tex
Update scripts: chap 8, appendix a.4, quotes, minor corrections, minor layout improvements

% --------------------------------------------------------------------------
\subsection{Prime number tests}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vskip +10 pt
\subsubsection{$f(n) = n^2 - 79 \cdot n + 1,601$}
\label{L-Polynomfunktion02-1601}
    This function delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
difference prime values (the same ones as delivered by the function in 
chapter~\ref{L-Polynomfunktion01-41}).
$$
\begin{array}{|ll||ll|}
\hline
f(0) = 1,601    & \mapsto {\rm ~prime} &  f(28) = 173    & \mapsto {\rm ~prime}
\\
f(1) = 1,523    & \mapsto {\rm ~prime} &  f(29) = 151    & \mapsto {\rm ~prime}
\\
f(2) = 1,447    & \mapsto {\rm ~prime} &  f(30) = 131 & \mapsto {\rm ~prime} \\
f(3) = 1,373    & \mapsto {\rm ~prime} &  f(31) = 113 & \mapsto {\rm ~prime} \\
f(4) = 1,301    & \mapsto {\rm ~prime} &  f(32) = 97 & \mapsto {\rm ~prime} \\
f(5) = 1,231    & \mapsto {\rm ~prime} &  f(33) = 83 & \mapsto {\rm ~prime} \\
f(6) = 1,163    & \mapsto {\rm ~prime} &  f(34) = 71 & \mapsto {\rm ~prime} \\
f(7) = 1,097    & \mapsto {\rm ~prime} &  f(35) = 61 & \mapsto {\rm ~prime} \\
f(8) = 1,033    & \mapsto {\rm ~prime} &  f(36) = 53 & \mapsto {\rm ~prime} \\
f(9) = 971  & \mapsto {\rm ~prime} &  f(37) = 47 & \mapsto {\rm ~prime} \\
f(10) = 911 & \mapsto {\rm ~prime} &  f(38) = 43 & \mapsto {\rm ~prime} \\
f(11) = 853 & \mapsto {\rm ~prime} &  f(39) = 41 & \mapsto {\rm ~prime} \\
f(12) = 797 & \mapsto {\rm ~prime} &  f(40) = 41 & \mapsto {\rm ~prime} \\               
f(13) = 743 & \mapsto {\rm ~prime} &  f(41) = 43 & \mapsto {\rm ~prime} \\               
f(14) = 691 & \mapsto {\rm ~prime} &  f(42) = 47 & \mapsto {\rm ~prime} \\               
f(15) = 641 & \mapsto {\rm ~prime} &  f(43) = 53 & \mapsto {\rm ~prime} \\               
f(16) = 593 & \mapsto {\rm ~prime} &  \cdots  &  \\                                      
f(17) = 547 & \mapsto {\rm ~prime} &  f(77) = 1,447  & \mapsto {\rm ~prime} \\           
f(18) = 503 & \mapsto {\rm ~prime} &  f(78) = 1,523  & \mapsto {\rm ~prime} \\           
f(19) = 461 & \mapsto {\rm ~prime} &  f(79) = 1,601  & \mapsto {\rm ~prime} \\           
f(20) = 421 & \mapsto {\rm ~prime} &  f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prime!} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prime!} \\ 
f(22) = 347 & \mapsto {\rm ~prime} &  f(82) = 1,847  & \mapsto {\rm ~prime} \\           
f(21) = 383 & \mapsto {\rm ~prime} &  f(83) = 1,933  & \mapsto {\rm ~prime} \\           
f(22) = 347 & \mapsto {\rm ~prime} &  f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prime!} \\
f(23) = 313 & \mapsto {\rm ~prime} & & \\
f(24) = 281 & \mapsto {\rm ~prime} & & \\
f(25) = 251 & \mapsto {\rm ~prime} & & \\
f(26) = 223 & \mapsto {\rm ~prime} & & \\
f(27) = 197 & \mapsto {\rm ~prime} & & \\
\hline
\end{array} 
$$

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century B.C., he found an
extremely easy, automatic way of finding this out. To begin with, you write
down all numbers from 2 to $x$, circle 2, then cross out all multiples of 2.
Next, you circle the lowest number that hasn't been circled or crossed out
(3) and again cross out all multiples of this number, etc. You only need to
continue until you reach the largest number whose square is less than or
equal to $x$.\index{NT, Learning Tool for Number Theory}%
\footnote{%
    With the educational tool for number theory {\bf NT} you can apply the
    sieve of Eratosthenes in a computer-aided and guided way: Enter you own
    number and do the sieving step by step: See learning unit 1.2,
    pages 6/21 and 7/21.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
     \textbackslash{} Learning Tool for Number Theory}.
}

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.cs.berkeley.edu/~aaronson/prime.ps}
         {\texttt{http://www.cs.berkeley.edu/\~{}aaronson/prime.ps}}. \\
    Only after I had completed this article, did I come across the 
    extremely well-written paper by Scott Aaronson, which also offers
    a didactically well done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

@@

I would like to take this opportunity to thank Mr.\ Henrik Koy and Mr.\ Roger
Oyono for their very constructive proof-reading of this article.

>>>>

% --------------------------------------------------------------------------
\subsection[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vskip +10 pt
\subsubsection{$f(n) = n^2 - 79 \cdot n + 1,601$}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{PrimArith1}{Appendix A of this chapter} for the source code
   to compute the table using Pari-GP.\index{Pari-GP}
}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{PrimArith2}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NICHT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NICHT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NICHT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century B.C., he found an
extremely easy, automatic way of finding this out. To begin with, you write
down all numbers from 2 to $x$, circle 2, then cross out all multiples of 2.
Next, you circle the lowest number that hasn't been circled or crossed out
(3) and again cross out all multiples of this number, etc. You only need to
continue until you reach the largest number whose square is less than or
equal to $x$.%
\footnote{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    sieve of Eratosthenes in a computer-aided and guided way: Enter you own
    number and do the sieving step by step:
    See learning unit 1.2, pages 6/21 and 7/21.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.scottaaronson.com/writings/prime.pdf}
 {\texttt{http://www.scottaaronson.com/writings/prime.pdf}}. \\
    After I had completed this article, I did come across the 
    fine paper by Scott Aaronson, which also offers
    a didactically very well-done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

@@

I would like to take this opportunity to thank Mr.\ Henrik Koy and Mr.\ Roger
Oyono for their very constructive proof-reading of the first versions
of this article.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{}
\subsection*{Appendix A: Examples using Pari-GP}
\addcontentsline{toc}{subsection}{Appendix A: Examples using Pari-GP}
\index{Pari-GP}

This appendix contains the source code for Pari-GP to compute e.g. the
quadratic sample sequences for primes.

\hypertarget{PrimArith1}{}
\subsubsection*{Validation of the results of a quadratic ``prime number formula''}

Sequences like these in the function value table on page \pageref{SrcPrimArith1}
can be computed in Pari-GP with the following commands:

\vspace{-2ex}
\begin{verbatim}
print("N -- N^2 - 9*N + 61");
for(N=0,50, X=N^2 - 9*N + 61;
  if (isprime(X)==1,
    print(N, " -- ", X, "  is prime"),
    print(N, " -- ", X, "  is NOT prime")
     );
   );
\end{verbatim}

The output looks like this:
\vspace{-2ex}
\begin{verbatim}
N -- N^2 - 9*N + 61
0 -- 61  is prime
1 -- 53  is prime
2 -- 47  is prime
3 -- 43  is prime
4 -- 41  is prime
5 -- 41  is prime
6 -- 43  is prime
7 -- 47  is prime
8 -- 53  is prime
9 -- 61  is prime
10 -- 71  is prime
11 -- 83  is prime
12 -- 97  is prime
13 -- 113  is prime
14 -- 131  is prime
15 -- 151  is prime
16 -- 173  is prime
17 -- 197  is prime
18 -- 223  is prime
19 -- 251  is prime
20 -- 281  is prime
21 -- 313  is prime
22 -- 347  is prime
23 -- 383  is prime
24 -- 421  is prime
25 -- 461  is prime
26 -- 503  is prime
27 -- 547  is prime
28 -- 593  is prime
29 -- 641  is prime
30 -- 691  is prime
31 -- 743  is prime
32 -- 797  is prime
33 -- 853  is prime
34 -- 911  is prime
35 -- 971  is prime
36 -- 1033  is prime
37 -- 1097  is prime
38 -- 1163  is prime
39 -- 1231  is prime
40 -- 1301  is prime
41 -- 1373  is prime
42 -- 1447  is prime
43 -- 1523  is prime
44 -- 1601  is prime
45 -- 1681  is NOT prime
46 -- 1763  is NOT prime
47 -- 1847  is prime
48 -- 1933  is prime
49 -- 2021  is NOT prime
50 -- 2111  is prime
\end{verbatim}

\hypertarget{PrimArith2}{}
\subsubsection*{Percentage of primes, delivered by a quadratic ``prime number formula''}

Pari-GP command:
\vspace{-2ex}
\begin{verbatim}
A=0;Beg=0;End=1000;
for(N=Beg,End, X=N^2-9*N+61;
  if(isprime(X)==1, 
  print(N, " -- ", X, "  is prime");A++,
  print(N, " -- ", X, "  is NOT prime")
    ) );
print("Number of primes: ", A, ";  Percentage of primes: " (A*100)/((End-Beg)+1)*1.);
\end{verbatim}
The output looks like this:
\vspace{-2ex}
\begin{verbatim}
...
1000 -- 991061  is NOT prime
Number of primes: 584;  Percentage of primes: 58.34165834165834165834165834
\end{verbatim}
